---
title: "01_analyze_pain_scores.Rmd"
output: html_document
date: "2025-12-12"
---

```{r setup, include = FALSE}

library(readxl)
library(ggplot2)
library(dplyr)
library(scales)
library(meta)

```

######################################## LOAD DATA AND APPLY EXCLUSIONS ########################################

### Load data

```{r}

study_characteristics <- read_excel("meta-analysis data.xlsx", sheet = 1)

pain_scores <- read_excel("meta-analysis data.xlsx", sheet = 2) %>%
  
  # Rename iud insertion to iud placement
  mutate(`Procedure step` = ifelse(`Procedure step` == "IUD insertion", "IUD placement", `Procedure step`)) %>%
  
  # Rename afterwards to post-procedure
  mutate(`Procedure step` = ifelse(`Procedure step` == "Afterwards", "Post-procedure", `Procedure step`)) %>%
  
  # Rename studies to include "et al"
  mutate(Study = case_when(
    Study == "Mody 2012" ~ "Mody et al, 2012",
    Study == "Cirik 2013" ~ "Cirik et al, 2013",
    Study == "Karasu 2017" ~ "Karasu et al, 2017",
    Study == "Akers 2017" ~ "Akers et al, 2017",
    Study == "Dogan 2017" ~ "Dogan et al, 2017",
    Study == "Berens 2017" ~ "Berens et al, 2017",
    Study == "Mody 2018" ~ "Mody et al, 2018",
    Study == "Khedr 2024" ~ "Khedr et al, 2024"
  ))

```

### Exclude overall pain, tenaculum pain for mody 2018, and paracervical block pain for control arm (some studies measured a comparison pain at that timepoint)

```{r}

pain_scores <- pain_scores %>%
  
  # Drop overall pain scores (don't end up using it since only one study assessed that)
  filter(`Procedure step` != "Overall") %>%
  
  # Remove the tenaculum pain score for mody 2018 (since we only have it for the paracervical block group as the control group got a tenaculum site injection so not a useful comparison there)
  filter(!(Study == "Mody et al, 2018" & `Procedure step` == "Tenaculum placement")) %>%
  
  # Remove pain scores for paracervical block step for control arm
  filter(!(`Procedure step` == "Paracervical block" & Arm != "Paracervical block"))

```

######################################## VISUAL SUMMARY OF ALL PAIN RESULTS ########################################

### Pre-process pain scores for plotting

```{r}

pain_scores_for_plotting <- pain_scores %>%
  
  # Whether the average pain statistic is a mean or a median
  mutate(which_pain_statistic = case_when(
    Study %in% c("Mody et al, 2012", "Karasu et al, 2017", "Dogan et al, 2017", "Berens et al, 2017", "Khedr et al, 2024") ~ "mean",
    Study %in% c("Cirik et al, 2013", "Akers et al, 2017", "Mody et al, 2018") ~ "median"
  )) %>%
  
  # Whether patients are nulliparous, parous, or any
  # Note to self: one of the studies is listed as "any (mostly parous)"
  left_join(select(study_characteristics, Study, Parity), by = c("Study")) %>%
  
  # Collapse all control arms (no treatment, sham, placebo) into one control arm label
  mutate(Arm = ifelse(Arm == "Paracervical block", "Paracervical block", "Control")) %>%
  mutate(Arm = factor(Arm, levels = c("Paracervical block", "Control"))) %>%

  # Order procedure steps
  mutate(`Procedure step` = factor(`Procedure step`, levels = c("Speculum insertion", "Paracervical block", "Tenaculum placement", "Uterine sounding", "IUD placement", "Post-procedure"))) %>%
  
  # For pain scores on a scale of 0-100, scale the averages (means and medians) to 0-10
  mutate(Mean = ifelse(`Pain scale` == "0-100", Mean / 10, Mean),
         Median = ifelse(`Pain scale` == "0-100", Median / 10, Median)) %>%
    
  # Pull out just the statistics needed for plotting, since some rescaling was done (but not to all columns in this section), to be extra careful this does not accidentally get weirdly propagated throughout any later code
  select(Study, `Procedure step`, Arm, Mean, Median, which_pain_statistic, Parity)

```

### Plot all pain scores together

```{r}

# All points
ggplot() +
  geom_point(data = filter(pain_scores_for_plotting, which_pain_statistic == "mean"),
             aes(`Procedure step`, Mean, color = Arm, shape = which_pain_statistic)) +
  geom_point(data = filter(pain_scores_for_plotting, which_pain_statistic == "median"),
             aes(`Procedure step`, Median, color = Arm, shape = which_pain_statistic)) +
  scale_x_discrete(labels = label_wrap(10)) +
  scale_y_continuous(limits = c(0,10), n.breaks = 10, minor_breaks = c(0,10)) +
  labs(shape = "Type of average") +
  scale_shape_manual(values = c("circle", "square"), labels = c("Mean", "Median")) +
  scale_color_manual(values = c("Paracervical block" = "blue", "Control" = "red")) +
  ylab("Average pain score") +
  ggtitle("Pain scores across procedure steps") +
  theme_minimal()

```

### Plot study-specific pain scores

```{r}

# Make a function for plotting a study's pain scores
plot_study_pain_scores <- function(all_data, study_name) {
  
  # Subset to that study's data
  study_data <- filter(all_data, Study == study_name)
  
  # Pull that study's pain statistic
  study_pain_statistic <- unique(study_data$which_pain_statistic)
  
  # Pull whether paracervical block pain was measured before tenaculum placement pain
  study_pb_before_tenaculum <- unique(study_data$pb_before_tenaculum)
  
  # If the paracervical block pain was *not* measured before tenaculum placement pain (ie, what happened was: first injection at tenaculum site, then tenaculum placement, then the main 2 paracervical block injections), update order of procedure steps
  if (study_pb_before_tenaculum == FALSE) {
    study_data <- study_data %>%
        mutate(`Procedure step` = factor(`Procedure step`, levels = c("Speculum insertion", "Tenaculum placement", "Paracervical block", "Uterine sounding", "IUD placement", "Post-procedure")))
  }
  
  # If the study reported means, plot means
  if (study_pain_statistic == "mean") {
    study_plot <- ggplot() +
      geom_point(data = study_data, aes(`Procedure step`, Mean, color = Arm, shape = which_pain_statistic)) +
      scale_x_discrete(labels = label_wrap(10)) +
      scale_y_continuous(limits = c(0,10), n.breaks = 10, minor_breaks = c(0,10)) +
      labs(shape = "Type of average") +
      scale_shape_manual(values = c("circle"), labels = c("Mean")) +
      scale_color_manual(values = c("Paracervical block" = "blue", "Control" = "red")) +
      ylab("Average pain score") +
      ggtitle(paste(study_name)) +
      theme_minimal()
    
  # If the study reported medians, plot medians
  } else if (study_pain_statistic == "median") {
    study_plot <- ggplot() +
      geom_point(data = study_data, aes(`Procedure step`, Median, color = Arm, shape = which_pain_statistic)) +
      scale_x_discrete(labels = label_wrap(10)) +
      scale_y_continuous(limits = c(0,10), n.breaks = 10, minor_breaks = c(0,10)) +
      labs(shape = "Type of average") +
      scale_shape_manual(values = c("square"), labels = c("Median")) +
      scale_color_manual(values = c("Paracervical block" = "blue", "Control" = "red")) +
      ylab("Average pain score") +
      ggtitle(paste(study_name)) +
      theme_minimal()
  }
  
  return(study_plot)
  
}

```

### Plot each study's pain scores

```{r}

# Specify order of whether paracervical block pain was measured before tenaculum placement pain
pain_scores_with_step_order <- pain_scores_for_plotting %>%
  mutate(pb_before_tenaculum = case_when(
    # In these five studies, paracervical block pain was measured after tenaculum placement pain
    Study %in% c("Mody et al, 2012", "Akers et al, 2017", "Berens et al, 2017", "Mody et al, 2018", "Khedr et al, 2024") ~ FALSE,
    # Of the remaining three studies, two measured paracervical block pain before tenaculum placement pain and one used ring forceps instead of a tenaculum
    TRUE ~ TRUE
  ))

# For each study, plot pain
for (study_name in unique(pain_scores_with_step_order$Study)) {
  print(plot_study_pain_scores(pain_scores_with_step_order, study_name))
}

```

######################################## META-ANALYSIS ########################################

### Pre-process pain scores for meta-analysis

```{r}

# Akers 2017 reported medians and confidence intervals
# So for the meta-analysis, we don't have enough info to approximate means using the standard methods (ie using median and iqr or range) using the built-in functionality in the meta package
# Instead, we will substitute the medians for means and then estimate standard deviation using the confidence intervals
pain_scores_with_updated_akers <- pain_scores %>%
  left_join(select(study_characteristics, Study, `Participants in paracervical block arm`, `Participants in control arm`), by = c("Study" = "Study")) %>%
  dplyr::rename(pb_n = `Participants in paracervical block arm`,
                c_n = `Participants in control arm`) %>%
  mutate(Mean = ifelse(Study == "Akers et al, 2017", Median, Mean),
         `SD of mean` = case_when(
           # Paracervical block arm
           Study == "Akers et al, 2017" & Arm == "Paracervical block" ~
             # Calculation for paracervical block arm:
             # The quantity [CI - mean] is approximated by calculating half the CI
             # This is then divided by the confidence level (here, 1.96)
             # Then, we multiply by the square root of the sample size (of paracervical block arm)
             sqrt(pb_n) * ( ( (`Upper 95 CI of median` - `Lower 95 CI of median`) / 2 ) / 1.96),
           # Control arm
           Study == "Akers et al, 2017" & Arm != "Paracervical block" ~
             # Analogous calculation for the control arm (using sample size of control arm)
             sqrt(c_n) * ( ( (`Upper 95 CI of median` - `Lower 95 CI of median`) / 2 ) / 1.96),
           TRUE ~ `SD of mean`))

# Make separate columns for paracervical block and control arms
pb_pain_scores <- pain_scores_with_updated_akers %>%
  filter(Arm == "Paracervical block") %>%
  mutate(pb_mean = Mean,
         pb_mean_sd = `SD of mean`,
         pb_median = Median,
         pb_q1 = Q1,
         pb_q3 = Q3,
         pb_min = Min,
         pb_max = Max) %>%
  select(c(1:3), c(18:26))
c_pain_scores <- pain_scores_with_updated_akers %>%
  filter(Arm != "Paracervical block") %>%
  mutate(c_mean = Mean,
         c_mean_sd = `SD of mean`,
         c_median = Median,
         c_q1 = Q1,
         c_q3 = Q3,
         c_min = Min,
         c_max = Max) %>%
  # Note to self: the difference in the column numbers here vs in analogous lines above is not a bug - some redundant info is already captured in the above lines and does not need to be recaptured here
  select(c(1,3), c(20:26))

# Combine
pain_scores_combined <- left_join(pb_pain_scores, c_pain_scores, by = c("Study" = "Study", "Procedure step" = "Procedure step")) %>%
  
  # Filter for steps that we are performing meta-analysis on
  filter(`Procedure step` %in% c("Tenaculum placement", "Uterine sounding", "IUD placement", "Post-procedure"))

# For studies that measured pain on a 0-100 scale, rescale all pain columns to 0-10 scale
pain_scores_0_10_scale <- filter(pain_scores_combined, `Pain scale` == "0-10")
pain_scores_0_100_scale_rescaled_to_0_10_scale <- filter(pain_scores_combined, `Pain scale` == "0-100") %>%
  mutate(across(6:19, ~ .x / 10))
pain_scores_for_meta_analysis <- rbind(pain_scores_0_10_scale, pain_scores_0_100_scale_rescaled_to_0_10_scale) %>%
  
  # Order trials by year
  mutate(Study = factor(Study, levels = c("Mody et al, 2012", "Cirik et al, 2013", "Karasu et al, 2017", "Akers et al, 2017", "Dogan et al, 2017", "Berens et al, 2017", "Mody et al, 2018", "Khedr et al, 2024"))) %>%
  arrange(Study)

```

### Tenaculum placement meta-analysis

```{r}

# Meta-analysis
tenaculum_placement_meta_analysis_results <- metacont(
  # Experimental arm is paracervical block
  n.e = pb_n, mean.e = pb_mean, sd.e = pb_mean_sd, median.e = pb_median, q1.e = pb_q1, q3.e = pb_q3, min.e = pb_min, max.e = pb_max,
  # Control arm is no treatment, sham, or placebo
  n.c = c_n, mean.c = c_mean, sd.c = c_mean_sd, median.c = c_median, q1.c = c_q1, q3.c = c_q3, min.c = c_min, max.c = c_max,
  # Label studies, arms, pain directions
  studlab = Study,
  label.e = "Paracervical block", label.c = "Control",
  label.left = "Reduces pain", label.right = "Increases pain",
  # Data for analysis
  data = filter(pain_scores_for_meta_analysis,
                `Procedure step` == "Tenaculum placement"),
  # Summary measure is mean difference
  sm = "MD",
  # Performing meta-analysis under a random effects model
  random = TRUE)

# Forest plot
tenaculum_placement_meta_analysis_results
forest(tenaculum_placement_meta_analysis_results, col.inside = "black")

```

### Uterine sounding meta-analysis

```{r}

# Meta-analysis
uterine_sounding_meta_analysis_results <- metacont(
  # Experimental arm is paracervical block
  n.e = pb_n, mean.e = pb_mean, sd.e = pb_mean_sd, median.e = pb_median, q1.e = pb_q1, q3.e = pb_q3, min.e = pb_min, max.e = pb_max,
  # Control arm is no treatment, sham, or placebo
  n.c = c_n, mean.c = c_mean, sd.c = c_mean_sd, median.c = c_median, q1.c = c_q1, q3.c = c_q3, min.c = c_min, max.c = c_max,
  # Label studies, arms, pain directions
  studlab = Study,
  label.e = "Paracervical block", label.c = "Control",
  label.left = "Reduces pain", label.right = "Increases pain",
  # Data for analysis
  data = filter(pain_scores_for_meta_analysis,
                `Procedure step` == "Uterine sounding"),
  # Summary measure is mean difference
  sm = "MD",
  # Performing meta-analysis under a random effects model
  random = TRUE)

# Forest plot
uterine_sounding_meta_analysis_results
forest(uterine_sounding_meta_analysis_results, col.inside = "black")

```

### IUD placement meta-analysis

```{r}

# Meta-analysis
iud_placement_meta_analysis_results <- metacont(
  # Experimental arm is paracervical block
  n.e = pb_n, mean.e = pb_mean, sd.e = pb_mean_sd, median.e = pb_median, q1.e = pb_q1, q3.e = pb_q3, min.e = pb_min, max.e = pb_max,
  # Control arm is no treatment, sham, or placebo
  n.c = c_n, mean.c = c_mean, sd.c = c_mean_sd, median.c = c_median, q1.c = c_q1, q3.c = c_q3, min.c = c_min, max.c = c_max,
  # Label studies, arms, pain directions
  studlab = Study,
  label.e = "Paracervical block", label.c = "Control",
  label.left = "Reduces pain", label.right = "Increases pain",
  # Data for analysis
  data = filter(pain_scores_for_meta_analysis,
                `Procedure step` == "IUD placement"),
  # Summary measure is mean difference
  sm = "MD",
  # Performing meta-analysis under a random effects model
  random = TRUE)

# Forest plot
iud_placement_meta_analysis_results
forest(iud_placement_meta_analysis_results, col.inside = "black")

```

### Post-procedure meta-analysis

```{r}

# Meta-analysis
post_procedure_meta_analysis_results <- metacont(
  # Experimental arm is paracervical block
  n.e = pb_n, mean.e = pb_mean, sd.e = pb_mean_sd, median.e = pb_median, q1.e = pb_q1, q3.e = pb_q3, min.e = pb_min, max.e = pb_max,
  # Control arm is no treatment, sham, or placebo
  n.c = c_n, mean.c = c_mean, sd.c = c_mean_sd, median.c = c_median, q1.c = c_q1, q3.c = c_q3, min.c = c_min, max.c = c_max,
  # Label studies, arms, pain directions
  studlab = Study,
  label.e = "Paracervical block", label.c = "Control",
  label.left = "Reduces pain", label.right = "Increases pain",
  # Data for analysis
  data = filter(pain_scores_for_meta_analysis,
                `Procedure step` == "Post-procedure"),
  # Summary measure is mean difference
  sm = "MD",
  # Performing meta-analysis under a random effects model
  random = TRUE)

# Forest plot
post_procedure_meta_analysis_results
forest(post_procedure_meta_analysis_results, col.inside = "black")

```






